### 2013-05-31 branching-transit.tex
WorkingWiki importation

\item This induces a distribution on $T$, the time that a successful excursion took to cross:
assuming, as is asymptotically correct, that the number of arriving lineages is independent of how long it took to get there,
if $\tau_x$ is the random time of arrival of a successful lineage,
\begin{align}
 \P\{\tau_x \in dt\} = C_{s_d,x} \frac{(1-s_d)^t}{t^{3/2}} \exp\left(-\frac{x^2}{2t}\right) ,
\end{align}
where $C_{s_d,x}$ is a constant depending only on $s_d$ and $x$.

@@

\item Radial symmetry implies a big annulus of habitat around the central patch.
If instead there are two smallish patches in two dimensions, 
we need only multiply the one-dimensional case by the probability that the $y$-coordinate is within the required range,
which is something involving the error function, or for large $x$,
is the width of the patch multiplied by $1/(2 \pi \sigma^2 t)$.  
This replaces the $t^{3/2}$ by a $t^{5/2}$, and back to G\&R, we have that the Laplace transform is in terms of $K_{3/2}$ instead of $K_{1/2}$;
and this is found in G\&R 8.468, $K_{3/2}(z) = \sqrt{\pi/2z} e^{-z} (1+1/z)$.
The formula corresponding to equation \eqref{eqn:first_integral} above is then
\begin{align}
  \E[e^{-\ell \tau_x}] \propto \sqrt{\pi} \frac{2\sqrt{\ell+\alpha}}{x^2} e^{-x\sqrt{2(\ell+\alpha)}} \left( 1 + \frac{1}{x\sqrt{2(\ell+\alpha)}} \right) ,
\end{align}
and so
\begin{align}
  \E[e^{-\ell \tau_x}] &= \frac{\sqrt{\ell+\alpha}}{\sqrt{\alpha}} e^{-x\sqrt{2}\left(\sqrt{\ell+\alpha}-\sqrt{\alpha}\right)} \left( 1 + \frac{1}{x\sqrt{2(\ell+\alpha)}} \right) \left( 1 + \frac{1}{x\sqrt{2(\alpha)}} \right)^{-1} \\
  &= \frac{x\sqrt{2(\ell+\alpha)}+1}{x\sqrt{2\alpha}+1} e^{-x\sqrt{2}\left(\sqrt{\ell+\alpha}-\sqrt{\alpha}\right)} .
\end{align}

\item Differentiating the above, we get that $\E[\tau_x] = x^2 / (x\sqrt{2\alpha}+1) \sim x/\sqrt{2\alpha}$ and $\var[\tau_x] = x^3 / \sqrt{2\alpha}(x\sqrt{2\alpha}+1)^2 \sim x/(2\alpha)^{3/2}$; in fact, the same scaling produces the same limit law, since the $\ell/\sqrt{x}$ term dissappears.

@@

A more back-of-the-envelope calculation of this would be:
we know that the probability a branching process arrives at the patch is given by $(1/\mu_K)\exp(-x\sqrt{2 \alpha}/\sigma)$;
if it does it is present in $\mu_K$ copies, each of which leave a total of $1/s_m$ offspring.
To translate this to the above we would need the typical area covered by these (in some sense?).
% which must be around $\sigma \sqrt{2 \alpha}$.

@@

We can calculate the rate of ``core'' establishment, i.e.\ the mean number of reproductive events that lead to establishment as defined below, in two ways:
\begin{align}
  (\mbox{rate of establishment}) &= (\mbox{outflux of BPs from old patch}) \times (\mbox{\# estab events per BP}) \\
    \begin{split} &= (\mbox{outflux of BPs from old patch}) \times (\mbox{prob of estab of a distant BP}) \\
  & \qquad \qquad \times (\mbox{\# estab events per successful BP}) \end{split} \\
  \begin{split} &= (\mbox{subcritical occupation density}) \times \rho \times (\mbox{\# offspring per subcritical indiv}) \\
    & \qquad \qquad \times (\mbox{prob of estab in the patch}) . \end{split} 
\end{align}
The probability of establishment of a branching process is in equation~\eqref{eqn:bp_prob_estab};
the subcritical occupation density is the integral of equation~\eqref{eqn:eqfreq} over the patch;
the number of (extra) offspring per subcritical individual is $s_B+s_m$;
and the probability of establishment in the patch is $2s_B/\xi^2$.
The ``outflux of branching processes'' we have not computed;
but we have computed the occupation density from this branching process approximation above,
so equating ($\rho$ times $C\exp(-\sqrt{2 s_m}x/\sigma)$) and (outflux times equation~\eqref{eqn:bp_occupation})
we have that the effective outflux (from the edge of the already occupied patch!) is 
\begin{align}
  (\mbox{outflux of BPs from old patch}) = C \rho \sigma \sqrt{2 \alpha} ,
\end{align}
rather than $C\rho$ which we might have guessed.
The number of establishment events per BP is as follows, in averages: 
with prob $1/\mu_K \exp(-\sqrt{\alpha} x/ \sigma)$ a total of $\mu_K$ indivs arrive at the new patch; 
each produce $1/s_m$ total core offspring;
each of these produce $s_B+s_m$ extra offspring which each have chance $2s_B/\xi^2$ of establishing,
i.e.
\begin{align}
  (\mbox{\# estab events per BP}) = \frac{(s_B+s_m) 2 s_B }{ s_m \xi^2 } \exp\left(-\frac{\sqrt{\alpha} x}{\sigma}\right) .
\end{align}
\textbf{Now} note that what we \textbf{actually} want to compute is not this rate, but rather 
\begin{align}
  \migrate &= (\mbox{outflux of BPs from old patch}) \times (\mbox{prob of estab of a distant BP}) \\
  &= C \rho \sqrt{\alpha} \times \frac{2 s_b}{\sqrt{\pi} \xi^2} \exp\left(- \frac{\sqrt{\alpha}x}{\sigma} \right) .
\end{align}
Finally, combining these, we can also compute that
\begin{align}
  (\mbox{\# estab events per successful BP}) = \frac{s_B+s_m}{ \sqrt{s_m} } .
\end{align}

>>>>

\item This induces a distribution on $T$, the time that a successful excursion took to cross:
assuming, as is asymptotically correct, that the number of arriving lineages is independent of how long it took to get there,
if $\tau_x$ is the random time of arrival of a successful lineage,
\begin{align} \label{eqn:arrival_time_density}
 \P\{\tau_x \in dt\} = C_{s_d,x} \frac{(1-s_d)^t}{t^{3/2}} \exp\left(-\frac{x^2}{2t}\right) ,
\end{align}
where $C_{s_d,x}$ is a constant depending only on $s_d$ and $x$.

@@

\item Radial symmetry implies a big annulus of habitat around the central patch.
If instead there are two smallish patches in two dimensions, 
we need only multiply the one-dimensional case by the probability that the $y$-coordinate is within the required range,
which is something involving the error function, or for large $x$,
is the width of the patch multiplied by $1/\sqrt{2 \pi \sigma^2 t}$.  
This replaces the $t^{3/2}$ in \eqref{eqn:arrival_time_density} by $t^2$, and back to G\&R, we have that the Laplace transform is in terms of $K_{1}$ instead of $K_{1/2}$;
in this case, we have that
\begin{align}
    \P\{ \tau < \infty \} &= \frac{x dy}{2\pi \sigma^4} \int_0^\infty t^{-2} \exp\left\{-\frac{x^2}{2\sigma^2 t} - \alpha t \right\} dt \\
                          &= \frac{x dy}{2\pi \sigma^4} 2 \left( \frac{ x^2 }{ 2 \alpha \sigma^2 } \right)^{-1/2} K_1\left(2 \sqrt\frac{x^2 \alpha}{2\sigma^2} \right) \\
                          &= \frac{dy\sqrt{2\alpha}}{\pi \sigma^3} K_1\left(\frac{x \sqrt{2\alpha}}{\sigma} \right)  \qquad \text{in } d=2
\end{align}
and we get $\E[e^{-\ell \tau_x}]$ by replacing $\alpha$ with $\alpha+\ell$.

% WRONGLY BASED ON 1/t INSTEAD OF 1/\sqrt{t}:
% and this is found in G\&R 8.468, $K_{3/2}(z) = \sqrt{\pi/2z} e^{-z} (1+1/z)$.
% The formula corresponding to equation \eqref{eqn:first_integral} above is then
% \begin{align}
%   \E[e^{-\ell \tau_x}] \propto \sqrt{\pi} \frac{2\sqrt{\ell+\alpha}}{x^2} e^{-x\sqrt{2(\ell+\alpha)}} \left( 1 + \frac{1}{x\sqrt{2(\ell+\alpha)}} \right) ,
% \end{align}
% and so
% \begin{align}
%   \E[e^{-\ell \tau_x}] &= \frac{\sqrt{\ell+\alpha}}{\sqrt{\alpha}} e^{-x\sqrt{2}\left(\sqrt{\ell+\alpha}-\sqrt{\alpha}\right)} \left( 1 + \frac{1}{x\sqrt{2(\ell+\alpha)}} \right) \left( 1 + \frac{1}{x\sqrt{2(\alpha)}} \right)^{-1} \\
%   &= \frac{x\sqrt{2(\ell+\alpha)}+1}{x\sqrt{2\alpha}+1} e^{-x\sqrt{2}\left(\sqrt{\ell+\alpha}-\sqrt{\alpha}\right)} .
% \end{align}
% 
% \item Differentiating the above, we get that $\E[\tau_x] = x^2 / (x\sqrt{2\alpha}+1) \sim x/\sqrt{2\alpha}$ and $\var[\tau_x] = x^3 / \sqrt{2\alpha}(x\sqrt{2\alpha}+1)^2 \sim x/(2\alpha)^{3/2}$; in fact, the same scaling produces the same limit law, since the $\ell/\sqrt{x}$ term dissappears.
% 
\end{itemize}

@@

We can combine this with the formula \eqref{eqn:eqfreq} for the equilibrium frequency $q(r) \approx C \exp( -\sqrt{2 \alpha} r / \sigma)$
to estimate $\lambda$, the effective outflux from the occupied patch: equating $\rho q(r)$ with $\lambda$ multiplied by equation \eqref{eqn:bp_occupation}, we get
\begin{align} \label{eqn:outflux_from_patch}
    \lambda = \text{(outflux from old patch)} = \begin{cases}
        C \rho \sigma \sqrt{2\alpha} \quad & d=1 \\
    C \rho \sigma^{3/2} (2\alpha)^{1/4} \sqrt{2 \pi x} \quad & d=2 
    \end{cases}
\end{align}
Note that this differs from the value $C\rho$ which we might have guessed.

\emph{The following is not making sense to me at the moment.} A more back-of-the-envelope calculation of this would be:
we know that the probability a branching process arrives at the patch is given by $(1/\mu_K)\exp(-x\sqrt{2 \alpha}/\sigma)$;
if it does it is present in $\mu_K$ copies, each of which leave a total of $1/s_m$ offspring.
To translate this to the above we would need the typical area covered by these (in some sense?).
% which must be around $\sigma \sqrt{2 \alpha}$.

@@

From the main paper, we have that
\begin{align} \label{eqn:bp_prob_estab}
    \text{(prob of estab of a distant BP)} &\approx \frac{2s_b}{\xi^2} \exp\left( - \frac{\sqrt{2 \alpha} x}{\sigma} \right) ,
\end{align}
and the subcritical occupation density per unit area at distance $r$ is
\begin{align} \label{eqn:eqfreq}
    q(r) \approx C \exp( -\sqrt{2 \alpha} r / \sigma).
\end{align}

We can calculate the rate of ``core'' establishment, i.e.\ the mean number of reproductive events that lead to establishment as defined below, in two ways:
\begin{align}
    (\mbox{rate of establishment}) &= (\mbox{outflux of BPs from old patch}) \times (\mbox{\# estab events per BP}) \label{eqn:heuristic1a} \\
    \begin{split} &= (\mbox{subcritical occupation density}) \times \rho \times (\mbox{\# offspring per subcritical indiv}) \label{eqn:heuristic1b} \\
    & \qquad \qquad \times (\mbox{prob of estab in the patch}) . \end{split} 
\end{align}
The subcritical occupation density is the integral of equation~\eqref{eqn:eqfreq} over the patch;
the number of (extra) offspring per subcritical individual is $s_B+s_m$;
and the probability of establishment of each such extra offspring in the patch is $2s_B/\xi^2$.
The effective outflux of branching processes is computed in equation \eqref{eqn:outflux_from_patch}.
The number of establishment events per BP is as follows, in averages: 
with prob $1/\mu_K \exp(-\sqrt{\alpha} x/ \sigma)$ a total of $\mu_K$ indivs arrive at the new patch; 
each produce $1/s_m$ total core offspring;
each of these produce $s_B+s_m$ extra offspring which each have chance $2s_B/\xi^2$ of establishing,
i.e.
\begin{align}
  (\mbox{\# estab events per BP}) = \frac{(s_B+s_m) 2 s_B }{ s_m \xi^2 } \exp\left(-\frac{\sqrt{2\alpha} x}{\sigma}\right) .
\end{align}

However, this formulation counts separately each of possible multiple establishment events by a single transiting family.
Writing
\begin{align} \label{eqn:heuristic2}
  \begin{split} (\mbox{rate of establishment}) &= (\mbox{outflux of BPs from old patch}) \times (\mbox{prob of estab of a distant BP}) \\
  & \qquad \qquad \times (\mbox{\# estab events per successful BP}) \end{split},
\end{align}
for $\migrate$ we actually want to omit the last term.
Using the probability of establishment of a distant branching process is in equation~\eqref{eqn:bp_prob_estab}, this is
\begin{align}
  \migrate &= (\mbox{outflux of BPs from old patch}) \times (\mbox{prob of estab of a distant BP}) \\
           &= \begin{cases}
    C \rho \sqrt{2 \alpha} \times \frac{2 s_b}{\sqrt{\pi} \xi^2} \exp\left(- \frac{\sqrt{2\alpha}x}{\sigma} \right)  \qquad & d=1 \\
    C \rho \sqrt{2 \alpha} \times \frac{ 2 s_b \sqrt{2 \alpha A} }{ \xi^2 \pi \sigma^3} K_1\left(\frac{x\sqrt{2\alpha}}{\sigma}\right)  \qquad & d=2 .
\end{cases}
\end{align}
Finally, combining \eqref{eqn:heuristic1b} and \eqref{eqn:heuristic2}, we can also compute that
\begin{align}
  (\mbox{\# estab events per successful BP}) = \frac{s_B+s_m}{ \sqrt{s_m} } .
\end{align}

### 2013-05-31 patchy-selection-paper.tex
WorkingWiki importation

Due to migration--selection balance, 
there will be a small number of $B$ alleles descended from those of the first patch
present in the regions where they are disadvantageous;
the expected frequency at distance $r$ may again be expressed in terms of Jacobi elliptic functions \citep[Appendix XXX][]{barton},
although the equations are again not particularly helpful.
This expected frequency is the \emph{time-averaged} occupation frequency,
i.e.\ the average frequency of the allele at each location over a long time period,
below we study the fluctuations about this mean.
More helpful than the exact expressions,
we know that if $q(r)$ is the expected proportion present at migration--selection balance at distance $r$ from the first patch
in the absence of the second patch, 
then for large $r$ this proportion decays exponentially, 
\begin{align} \label{eqn:eqfreq}
  q(r) \approx C \exp( -\sqrt{2 s_m} r / \sigma),
\end{align}
where $C$ is a constant depending on the geographic shape of the populations and the selection coefficients. 
(Probably $C=1/2$ is close enough for applications; see Appendix XXX.)
\plr{same appendix with sims as above}
\gc{Should we also use this approx. to get the approx. rate of establishing mutations
arising outside of the patch, above. Feel like in the mutation rate
part we could pull $s_B$ out front of the rate as well.} 
\plr{huh?}

This allows us to get a good sense of the time scale of adaptation by migration between separated patches, as follows.
First imagine that the new allele is not advantageous in the second patch,
so that the expected number of alleles present in a small area $\Delta$ at distance $r$ from the already established patch
is $\Delta \rho q(r)$.
As these copies are disadvantageous, they form a subcritical branching process, which die out fairly soon.
Now add to this local selective advantage by stipulating that any copy of the allele in the new patch
has ``extra'' offspring that are marked in some way, on average $(1+s_B)-(1-s_m)=s_B+s_m$ additional offspring per generation.
If the probability of establishment \emph{in the new patch} is $p(x)$ as above,
then each of these new, marked copies appearing at location $x$ has probability $p(x)$ of establishing.
(This general relation is stated formally in Appendix XXX.)
Therefore, the total rate of establishment 
(i.e.\ the mean number of successfully establishing migrant lineages per generation)
is $\rho (s_B+s_m) \int q(|x|) p(x) dx$.
Using a similar approximation as in section \ref{ss:patchymutation},
we can approximate this, for a patch of area $A$ at distance $R$ from an already occupied patch, as
\begin{align}
  \migrate(R) = \rho A \frac{s_B}{\xi^2} (s_B+s_m) \exp\left( -\frac{ \sqrt{2 s_m} R}{\sigma} \right),
\end{align}
where we have set $C=1/2$ in equation \eqref{eqn:eqfreq}.
Using different methods, we arrive at the same functional form below in section \ref{ss:haplotype_length}.
(Note that if the size of the new patch is large relative to $\sqrt{s_m}/\sigma$,
then the integral of $q(x)$ over the patch will not be well approximated by $A \times q(R)$,
but the adjustment is easy to make.)

@@

The process $\Phi_t$ is an irreducible, aperiodic finite-state Markov chain with absorbing states at 0 and 1;
therefore, the inevitable outcome for finite $N$ is extinction of one type or another.
Setting the left-hand side to zero in \eqref{eqn:discrete_generator} gives us equations for the moments of the stationary distribution,
for instance, if $\phi(x) = \lim_{t \to \infty} \E[ \Phi_t(x) ]

>>>>

Due to migration--selection balance, 
there will be a small number of $B$ alleles descended from those of the first patch
present in the regions where they are disadvantageous;
the expected frequency at distance $r$ may again be expressed in terms of Jacobi elliptic functions \citep[Appendix XXX][]{barton},
although the equations are again not particularly helpful.
This expected frequency is the \emph{time-averaged} occupation frequency,
i.e.\ the average frequency of the allele at each location over a long time period,
below we study the fluctuations about this mean.
More helpful than the exact expressions,
we know that if $q(r)$ is the expected proportion present at migration--selection balance at distance $r$ from the first patch
in the absence of the second patch, 
then for large $r$ this proportion decays exponentially, 
\begin{align} \label{eqn:eqfreq}
  q(r) \approx C \exp( -\sqrt{2 s_m} r / \sigma),
\end{align}
where $C$ is a constant depending on the geographic shape of the populations and the selection coefficients. 
(Probably $C=1/2$ is close enough for applications; see Appendix XXX.)
\plr{same appendix with sims as above}
\gc{Should we also use this approx. to get the approx. rate of establishing mutations arising outside of the patch, above.} 
\plr{Difficulties with this... and I'm not sure how interesting that is?  Convince me otherwise?}
\gc{Feel like in the mutation rate part we could pull $s_B$ out front of the rate as well.} 
\plr{huh?}

As we show below in section \ref{ss:haplotype_length}, the rate of establishment, for a patch of area $A$ at distance $R$ from an already occupied patch, is approximately
\begin{align} \label{eqn:migrate}
    \migrate(R) = \rho \frac{s_B\sqrt{2s_m} }{\xi^2\sqrt{\pi}}  \exp\left( -\frac{ \sqrt{2 s_m} R}{\sigma} \right).
\end{align}
The exponential term is expected based on \eqref{eqn:eqfreq} -- 
since establishment requires rare long distant migrants to be present, and \eqref{eqn:eqfreq} gives the density of such --
but the constant requires other arguments.

% This allows us to get a good sense of the time scale of adaptation by migration between separated patches, as follows.
% First imagine that the new allele is not advantageous in the second patch,
% so that the expected number of alleles present in a small area $\Delta$ at distance $r$ from the already established patch
% is $\Delta \rho q(r)$.
% As these copies are disadvantageous, they form a subcritical branching process, which die out fairly soon.
% Now add to this local selective advantage by stipulating that any copy of the allele in the new patch
% has ``extra'' offspring that are marked in some way, on average $(1+s_B)-(1-s_m)=s_B+s_m$ additional offspring per generation.
% If the probability of establishment \emph{in the new patch} is $p(x)$ as above,
% then each of these new, marked copies appearing at location $x$ has probability $p(x)$ of establishing.
% (This general relation is stated formally in Appendix XXX.)
% Therefore, the total rate of establishment 
% (i.e.\ the mean number of successfully establishing migrant lineages per generation)
% is $\rho (s_B+s_m) \int q(|x|) p(x) dx$.
% Using a similar approximation as in section \ref{ss:patchymutation},
% we can approximate this, for a patch of area $A$ at distance $R$ from an already occupied patch, as
% \begin{align}
%   \migrate(R) = \rho A \frac{s_B}{\xi^2} (s_B+s_m) \exp\left( -\frac{ \sqrt{2 s_m} R}{\sigma} \right),
% \end{align}
% where we have set $C=1/2$ in equation \eqref{eqn:eqfreq}.
% Using different methods, we arrive at the same functional form below in section \ref{ss:haplotype_length}.
% (Note that if the size of the new patch is large relative to $\sqrt{s_m}/\sigma$,
% then the integral of $q(x)$ over the patch will not be well approximated by $A \times q(R)$,
% but the adjustment is easy to make.)

@@

The process $\Phi_t$ is an irreducible, aperiodic finite-state Markov chain with absorbing states at 0 and 1;
therefore, the inevitable outcome for finite $N$ is extinction of one type or another.
Setting the left-hand side to zero in \eqref{eqn:discrete_generator} gives us equations for the moments of the stationary distribution,
for instance, if $\phi(x) = \lim_{t \to \infty} \E[ \Phi_t(x) ]$
